Discover governing reactions from concentration data

### 2017-11-01 docs/manuscript.tex

	\section{The method}
	The underlying model that we want to fit the data to is a law of mass action type dynamical system. To this end, let $S$ be the number of species, then the concentration data at a time $t$ can be represented by a vector
	\[
	\mathbf{x}(t)=\begin{pmatrix}
	x_1(t)\\ \vdots \\ x_S(t)
	\end{pmatrix}\in \mathbb{R}^S.
	\]
	Further, one can choose $R$ possible ansatz reactions with their respective reaction function
	\[
	\textbf{y}_r(\textbf{x}(t))=\begin{pmatrix}
	y_{r,1}(\textbf{x}(t)) \\ \vdots \\ y_{r,S}(\textbf{x}(t))
	\end{pmatrix}
	\]
	so that the change of concentration for species $i$ at time $t_i$, is represented by the dynamical system
	\[
	\dot{\textbf{x}}_i(t) = \sum_{r=1}^{R}y_{r,i}(\textbf{x}(t))\xi_r,\quad i=1,\ldots, S,
	\]
	where $\xi_r$ are the to-be estimated macroscopic rates.

	When presented with a time series with $T$ observations, the data can be represented as a matrix
	\[
	\textbf{X} = \begin{pmatrix}
		x_1(t_1) & x_2(t_1) & \cdots & x_S(t_1) \\
		x_1(t_2) & x_2(t_2) & \cdots & x_S(t_2) \\
		\vdots   & \vdots   & \ddots & \vdots   \\
		x_1(t_T) & x_2(t_T) & \cdots & x_S(t_T)
	\end{pmatrix} \in \mathbb{R}^{T\times S}.
	\]
	Given this matrix, one can propose a library $\Theta(\textbf{X}) = \begin{pmatrix} \theta_1(\textbf{X}) & \theta_2(\textbf{X}) & \cdots & \theta_R(\textbf{X}) \end{pmatrix}$ of $R$ ansatz reactions with corresponding reaction functions
	\[
	\theta_r(\textbf{X}) = \begin{pmatrix}
	\textbf{y}_r(\textbf{X}_1)^T \\ \vdots \\ \textbf{y}_r(\textbf{X}_T)^T
	\end{pmatrix}\in \mathbb{R}^{T\times S},\quad r=1,\ldots,R,
	\]
	where $\textbf{X}_i$ denotes the $i$-th row in $X$. Applying the concentration trajectory to the library yields a data tensor $\Theta(\textbf{X})\in\mathbb{R}^{T\times S\times R}$. Following the approach of SINDy, the goal is now to find coefficients $\Xi = \begin{pmatrix} \xi_1 & \xi_2 & \cdots & \xi_R
	\end{pmatrix}^T$, so that
	\[
	\dot{\textbf{X}} = \Theta(\textbf{X})\Xi = \sum_{r=1}^{R}\theta_r(\textbf{X})\xi_r.
	\]
	In particular, the system is linear in the coefficients $\Xi$, which makes sparse regression tools such as LASSO \cite{Tibshirani1996, Hastie2009} applicable. To this end, one can consider the minimization problem to find $\hat{\Xi}$ such that
	\begin{align}
		\hat{\Xi} = \underset{\Xi}{\arg\min}\left( \frac{1}{2T}\left\| \dot{\textbf{X}} - \Theta(\textbf{X})\Xi \right\|_F^2 + \alpha\|\Xi\|_1 \right) \quad \text{subject to }\Xi \geq 0,
	\label{method:minimizationproblem}\end{align}
	where $\alpha\geq 0$ is a sparsity inducing hyperparameter. For $\alpha=0$ this problem reduces to constrained least-squares. For solving (\ref{method:minimizationproblem}) we apply the sequential least squares minimizer SLSQP, originally described in \cite{Kraft1988}, contained in, e.g., the software package SciPy \cite{SciPy}. As initial guess for the regulated problem we use the solution of (\ref{method:minimizationproblem}) with $\alpha=0$.

@@

	The problem now is to find a suitable $\alpha$ such that a parsimonious solution is picked. To this end, cross-validation is applied. The trajectories that were used as training and test data set are depicted in Figure \ref{ex2:test-train-concentration-curves}.

>>>>

	\section{The method}
	The underlying model that we want to fit the data to is a law of mass action type dynamical system. To this end, let $S$ be the number of species, then the concentration data at a time $t$ can be represented by a vector
	\begin{align}
	\mathbf{x}(t)=\begin{pmatrix}
	x_1(t)\\ \vdots \\ x_S(t)
	\end{pmatrix}\in \mathbb{R}^S.
	\end{align}
	Further, one can choose $R$ possible ansatz reactions with their respective reaction function
	\begin{align}
	\textbf{y}_r(\textbf{x}(t))=\begin{pmatrix}
	y_{r,1}(\textbf{x}(t)) \\ \vdots \\ y_{r,S}(\textbf{x}(t))
	\end{pmatrix}
	\end{align}
	so that the change of concentration for species $i$ at time $t_i$, is represented by the dynamical system
	\begin{align}
	\dot{\textbf{x}}_i(t) = \sum_{r=1}^{R}y_{r,i}(\textbf{x}(t))\xi_r,\quad i=1,\ldots, S,
	\end{align}
	where $\xi_r$ are the to-be estimated macroscopic rates.

	When presented with a time series with $T$ observations, the data can be represented as a matrix
	\begin{align}
	\textbf{X} = \begin{pmatrix}
		x_1(t_1) & x_2(t_1) & \cdots & x_S(t_1) \\
		x_1(t_2) & x_2(t_2) & \cdots & x_S(t_2) \\
		\vdots   & \vdots   & \ddots & \vdots   \\
		x_1(t_T) & x_2(t_T) & \cdots & x_S(t_T)
	\end{pmatrix} \in \mathbb{R}^{T\times S}.
	\end{align}
	Given this matrix, one can propose a library $\Theta(\textbf{X}) = \begin{pmatrix} \theta_1(\textbf{X}) & \theta_2(\textbf{X}) & \cdots & \theta_R(\textbf{X}) \end{pmatrix}$ of $R$ ansatz reactions with corresponding reaction functions
	\begin{align}
	\theta_r(\textbf{X}) = \begin{pmatrix}
	\textbf{y}_r(\textbf{X}_1)^T \\ \vdots \\ \textbf{y}_r(\textbf{X}_T)^T
	\end{pmatrix}\in \mathbb{R}^{T\times S},\quad r=1,\ldots,R,
	\end{align}
	where $\textbf{X}_i$ denotes the $i$-th row in $X$. Applying the concentration trajectory to the library yields a data tensor $\Theta(\textbf{X})\in\mathbb{R}^{T\times S\times R}$. Following the approach of SINDy, the goal is now to find coefficients $\Xi = \begin{pmatrix} \xi_1 & \xi_2 & \cdots & \xi_R
	\end{pmatrix}^T$, so that
	\begin{align}
	\dot{\textbf{X}} = \Theta(\textbf{X})\Xi = \sum_{r=1}^{R}\theta_r(\textbf{X})\xi_r.
	\end{align}
	In particular, the system is linear in the coefficients $\Xi$, which makes sparse regression tools such as LASSO \cite{Tibshirani1996, Hastie2009} applicable. To this end, one can consider the minimization problem to find $\hat{\Xi}$ such that
	\begin{align}
		\hat{\Xi} = \underset{\Xi}{\arg\min}\left( \frac{1}{2T}\left\| \dot{\textbf{X}} - \Theta(\textbf{X})\Xi \right\|_F^2 + \alpha\|\Xi\|_1 \right) \quad \text{subject to }\Xi \geq 0,
	\label{method:minimizationproblem}\end{align}
	where $\alpha\geq 0$ is a sparsity inducing hyperparameter and $\|\cdot\|_F$ denotes the Frobenius norm. For $\alpha=0$ this problem reduces to constrained least-squares. For solving (\ref{method:minimizationproblem}) we apply the sequential least squares minimizer SLSQP, originally described in \cite{Kraft1988}, contained in, e.g., the software package SciPy \cite{SciPy}. As initial guess for the regulated problem we use the solution of (\ref{method:minimizationproblem}) with $\alpha=0$.

@@

	The problem now is to find a suitable $\alpha$ such that a parsimonious solution is picked. To this end, cross-validation is applied. The trajectories that were used as training and test data set are depicted in Figure \ref{ex2:test-train-concentration-curves}. \textcolor{red}{Perhaps use several test trajectories with different initial conditions to get something like a confidence interval?} As an indicator to the estimation error we use the scoring function
	\begin{align}
		f_\alpha = \frac{1}{T}\|\mathbf{\dot{X}}_\text{test} - \Theta (\textbf{X}_\text{test}) \Xi_\text{train}\|_F^2.
		\label{ex2:score-function}
	\end{align}
	For a selection of values $\alpha\in [0, 100]$, the model parameters $\Xi$ were estimated and the scoring function (\ref{ex2:score-function}) evaluated, yielding a graph as in Figure \ref{ex2:score-graph}.

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/cv_score}
		\end{center}
		\caption{Score function (\ref{ex2:score-function}) evaluated for a range of $\alpha$ values.}
		\label{ex2:score-graph}
	\end{figure}

	One can see that the minimum is at about
	\begin{align}
		\hat{\alpha} = \underset{\alpha\in[0,100]}{\arg\min}f_\alpha\approx 10.
	\end{align}
	When evaluating the minimization problem (\ref{method:minimizationproblem}) with this hyperparameter and applying a cut-off of $10^{-8}$ to the yielded parameters, one obtains
